Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
Work on proof.

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. We first argue that $\I$ satisfies a $\Sigma$-formula
$\psi$ that is closed under function expansion with respect to $\DDD$, and
where $\psi$ entails $\varphi_0$.
Since $\I$ is a model of $\varphi_0$,
there exists a set of literals $\psi_0$ over the atoms occurring in $\varphi_0$
such that $\I$ satisfies $\psi_0$ and $\psi_0$ entails $\varphi_0$.
\ajr{todo}

Since there exists a model for $\psi$ that is closed under function expansion
with respect to $\DDD$, then since $\DDD$ is admissible, by definition there
exists a model for $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
which is $\varphi$.
\qed
\end{proofsketch}

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be
interpreted as $\sortint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\sortint}.\;\, \lite\bigl( x \leq 0, \con{s}(x)
\teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargsort{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: any model satisfying $\varphi_{\con{s}}[t/x]$
must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a model for $\psi \wedge \varphi_\con{s}$,
which means there exists a model for formula~(\ref{eq:ex-before}).\xend
\end{example}

>>>>

Second, we show that if $\conv( \varphi )$ is satisfied by a
$\extendsig{\Sigma}$-interpretation $\I$, then $\varphi$ is satisfied by
$\Sigma$-interpretation $\J$. Since $\varphi$ is in definitional form with
respect to the functions defined by $\DDD$, we know $\varphi$ is of the form
$\DDD \wedge \varphi_0$. 
%We first argue that $\I$ satisfies a $\Sigma$-formula
%$\psi$ that is closed under function expansion with respect to $\DDD$, and
%where $\psi$ entails $\varphi_0$.
To do so, 
we first construct a sequence of $\extendsig{\Sigma}$-literals $M_0, M_1, \ldots$ such that $\I$ satisfies each $M_i$,
and a sequence of mappings $C^{\con f}_0, C^{\con f}_1, \ldots$ from values $\fargsort{f}^\I$ 
to tuples of terms of type $\tau_1 \times \cdots \times \tau_n$ 
for each function $\con{f}$ of type $\tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in $\sfundefs{\Sigma}$.
Let $M_0$ be a set of literals over the atoms occurring in $\varphi_0$
such that $\I$ satisfies each literal in $M_0$ and $\bigwedge M_0$ entails $\conv( \varphi_0 )$.
For each $\con{f} \in \sfundefs{\Sigma}$, let $C^{\con{f}}_0$ be the empty mapping.
For each $i \geq 0$, 
let $X_i$ be the set of constraints of the form $\exists a : {\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t )$ occurring in $M_i$
and let $L_i$ be the others in $M_i$.
For each $\con{f} \in \sfundefs{\Sigma}$, let $C^{\con{f}}_{i+1}$ be a mapping containing $C^{\con{f}}_i$ and a minimal number of entries
such that for all $\exists a : {\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t ) \in X_i$,
there exists a $u \mapsto \vec s \in C^{\con{f}}_{i+1}$ such that $\vecfarg{\con{f}}^\I( u ) = \vec s^\I = \vec t^\I$.
Let $M_{i+1}$ be the set containing $M_i$ as well as

Let $M^\ast$ be the limit of this sequence, which we partition into sets $X^\ast$ and $L^\ast$ in the same way as above.
Let $\psi$ be $\bigwedge L^\ast$.
We know $\psi$ is a $\Sigma$-formula since \ajr{blah}
We know $\psi$ is closed under function expansion with respect to $\DDD$ since \ajr{blah}
Finally, we know $\psi$ entails $\varphi_0$ since \ajr{blah}
Since there exists a model for $\psi$ that is closed under function expansion
with respect to $\DDD$, then since $\DDD$ is admissible, by definition there
exists a model for $\psi \wedge \DDD$, which entails $\DDD \wedge \varphi_0$,
which is $\varphi$.
\qed
\end{proofsketch}

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be
interpreted as $\sortint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\sortint}.\;\, \lite\bigl( x \leq 0, \con{s}(x)
\teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargsort{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: any model satisfying $\varphi_{\con{s}}[t/x]$
must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$ \ajr{revise this sentence, not quite accurate}. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$,
which means there exists a $\Sigma$-interpretation satisfying formula~(\ref{eq:ex-before}).\xend
\end{example}
